### 2022-03-30 conclusion/conclusion.tex
summary

The work that I have done for my thesis
was cyclic: It consists of three phases (three stops as we described them in
Chapter \ref{ch:intro}), each of which has the same parts: theory (calculus and
algorithms), engineering (implementation) and optimization (heuristics). While
the calculi were mostly developed by Alexander Bentkamp
\cite{bbcw-21-lfho,bbtvw-21-sup-lam,bbtv-21-full-ho-sup}, the task of developing
algorithms for indexing or unification, as well as the tasks of implementing and
optimizing the implementation were mainly mine.

Structuring the project cyclically turned out to be a great idea of my
supervisor. When I first started doing the research which will later form the
basis of Chapter \ref{ch:ehoh}, I was overwhelmed by the scope of the project
and felt uncertain if E could ever fully support $\lambda$-free higher-order
logic. Many of the extensions described in the  later chapters were even larger
in scope. However, as each extension relied on the previous one, I quickly
developed a feeling of where the possible pitfalls are and how to avoid them.
This gave me an invaluable source of encouragement, without which I could hardly
arrive at the end of the project.

@@

Even though \lsup{} became competitive, its implementation in Zipperposition
lagged behind the competition. To investigate why this is the case,
we manually analyzed hundreds of benchmarks which the competition proves, but
on which Zipperposition fails. This analysis forms the basis of work that is
described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques}. The main
conclusion was that the competition, most notably tableaux-based Satallax,
reasons better with formulas. Tableaux inferences are more intuitive than the
ones that superposition or resolution provers perform and are more in line with
how a mathematician structures a proof. Fitting these kind of
inferences into the superposition context was challenging, but it proved very
successful. After they were implemented in Zipperposition, it took its first
victory at higher-order division of CASC.

Our extension of \lsup{} was guided by performance, rather than completeness
with respect to full higher-order logic. After the first victory, we were left
wondering what were the kind of problems that our implementation cannot
solve, and started looking for complete full higher-order calculus. At the end,
we designed \osup{} which gave us the precise answer in the form of problems that
cannot be solved and the rules which are necessary to solve them.

While preparing for the following year's CASC competition, we found problems
that occur in practice for which the rules of \osup{} are necessary. However,
these problems occur in less than one percent of the whole TPTP library.
Furthermore, some of \osup{}'s rule are so explosive that they are disabled in
most of CASC portfolio configurations.

This lead us to take a radically different approach when extending Ehoh to
\ehohii{}. We envisioned \ehohii{} as a prover that excels on problems coming
from proof assistants. Unlike in TPTP, on these benchmarks hard, hand-crafted mathematical puzzles rarely occur.
Thus, we conjectured that extending \lfsup{} with the most rudimentary features
of $\lambda$ superposition such as full higher-order unification and
$\beta$-reduction-aware term orders will help us prove most of higher-order
problems without the explosion of complete approaches.

While this proved mostly true, we quickly realized that \ehohii{} should
assimilate most of incomplete Boolean reasoning techniques described in Chapters
\ref{ch:bools} and \ref{ch:ho-techniques}. All of them were straightforward to
implement except for dynamic clausification. It is possible that support for
this technique does not require profound changes to E(hoh)'s formula treatment,
but due to the lack of time we did not explore this alley.

This highly pragmatic approach proved successful not only on proof assistant
benchmarks: \ehohii{} also excels on TPTP benchmarks, trailing slightly behind
Zipperposition. By manually expecting the TPTP benchmarks that are out of reach
for \ehohii{}, but within the reach of Zipperposition, we realized that lack of
dynamic clausification might be the weight necessary to tip the scale in the
favor of \ehohii{} on TPTP benchmarks.

This extension of Ehoh to \ehohii{} finished the third cycle. Inspired by our
approach of extending techniques designed for weaker logics to work in the
context of richer logics, we decided to look for ways in which superposition can
assimilate the most successful techniques of SAT solving. Extension of hidden
literals was straightforward, but finding the right way to tame their infinite
nature in the first-order case, as well as the right way to integrate equality
in their definition was challenging. Predicate elimination posed challenges in
terms of finding the exact conditions in which it can be integrated with the
saturation loop. We first observed that blocked clause elimination destroys the
completeness of superposition calculus when we noticed that Zipperposition does
not prove some problems when the technique is enabled. After sifting through tens of
benchmarks and carefully examining the debugging information we realized that
this was not due to a bug in the implementation but due to incompatibility of
blocked clause elimination with redundancy criterion. In the end, we were left
somewhat disappointed to learn that SAT techniques do not scale good enough to be used
as superposition simplification techniques. However, we learned that they are
very successful as preprocessing techniques.

>>>>

The work described in this thesis
was cyclic: It consists of three phases (three stops as described in
Chapter \ref{ch:intro}), each of which has the same parts: theory (calculus and
algorithms), engineering (implementation) and optimization (heuristics). While
the calculi were mostly developed by Alexander Bentkamp
\cite{bbcw-21-lfho,bbtvw-21-sup-lam,bbtv-21-full-ho-sup}, the task of developing
algorithms for indexing or unification, as well as the tasks of implementing and
optimizing the implementation were mainly mine.

Structuring the project cyclically turned out to be a great idea of Jasmin
Blanchette. When I first started doing the research which will later form the
basis of Chapter \ref{ch:ehoh}, I was overwhelmed by the scope of the project
and felt uncertain if E could ever fully support $\lambda$-free higher-order
logic. Many of the extensions described in the later chapters were even larger
in scope. However, as each extension relied on the previous one, I quickly
developed a feeling of where the possible pitfalls are and how to avoid them.
This gave me an invaluable source of encouragement, without which I could hardly
arrive at the end of the project.

@@

Even though \lsup{} became competitive, its implementation in Zipperposition
lagged behind the competition. To investigate why this is the case,
we manually analyzed hundreds of benchmarks which the competition proves, but
on which Zipperposition fails. This analysis forms the basis of work that is
described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques}. The main
conclusion was that the competition, most notably the tableaux-based Satallax,
reasons better with formulas. Tableaux inferences are more intuitive than the
ones that superposition or resolution provers perform and are more in line with
how a mathematician structures a proof. Fitting these kind of
inferences into the superposition context was challenging, but it proved very
successful. After they were implemented in Zipperposition, it took its first
victory at higher-order division of CASC.

After its first victory, Zipperposition was successfully integrated in
Sledgehammer and it is now part of its default installation. We also noticed
that problems which only Zipperposition can solve do pop up from time to time in
practice. Antoine Defourné also integrated Zipperposition in TLA+ proof
assistant and observed that it helps prove some problems that were previously
out of reach \cite{ad-21-ho-tlaplus}.

Our extension of \lsup{} was guided by performance, rather than completeness
with respect to full higher-order logic. After the first victory, we were left
wondering what were the kind of problems that our implementation cannot
solve, and started looking for complete full higher-order calculus. At the end,
we designed \osup{}, which gave us the precise answer in the form of problems that
cannot be solved and the rules which are necessary to solve them.

While preparing for the following year's CASC competition, we found problems
that occur in practice for which the rules of \osup{} are necessary. However,
these problems occur in less than one percent of the whole TPTP library.
Furthermore, some of \osup{}'s rule are so explosive that they are disabled in
most CASC portfolio configurations.

This led us to take a radically different approach when extending Ehoh to
\ehohii{}. We envisioned \ehohii{} as a prover that excels on problems coming
from proof assistants. Unlike in TPTP, on these benchmarks hard, hand-crafted mathematical puzzles rarely occur.
Thus, we conjectured that extending \lfsup{} with the most rudimentary features
of $\lambda$-superposition such as full higher-order unification and
$\beta$-reduction-aware term orders will help us prove most of higher-order
problems without the explosion of complete approaches.

While this proved mostly true, we quickly realized that \ehohii{} should
assimilate most of incomplete Boolean reasoning techniques described in Chapters
\ref{ch:bools} and \ref{ch:ho-techniques}. All of them were straightforward to
implement except for dynamic clausification, which we did not implement. It is
possible that support for this technique does not require profound changes to
E(hoh)'s formula treatment, but due to the lack of time we did not explore this
alley.

This highly pragmatic approach proved successful not only on proof assistant
benchmarks: \ehohii{} also excels on TPTP benchmarks, trailing slightly behind
Zipperposition. In the coming months we plan to replace Ehoh by \ehohii{} in Sledgehammer.
As Ehoh is already one of the most successful Sledgehammer backends,
this will further improve the efficiency of hundreds of mathematicians and computer scientists.
By manually expecting the TPTP benchmarks that are out of reach
for \ehohii{}, but within the reach of Zipperposition, we realized that lack of
dynamic clausification might be the weight necessary to tip the scale in
favor of \ehohii{} on TPTP benchmarks.

This extension of Ehoh to \ehohii{} finished the third cycle. Inspired by our
approach of extending techniques designed for weaker logics to work in the
context of richer logics, we decided to look for ways in which superposition can
assimilate the most successful techniques of SAT solving. Extension of hidden
literals was straightforward, but finding the right way to tame their infinite
nature in the first-order case, as well as the right way to integrate equality
in their definition was challenging. Predicate elimination posed challenges in
terms of finding the exact conditions under which it can be integrated with the
saturation loop. We first observed that blocked clause elimination destroys the
completeness of superposition calculus when we noticed that Zipperposition does
not prove some problems when the technique is enabled. After sifting through tens of
benchmarks and carefully examining the debugging information, we realized that
this was not due to a bug in the implementation but due to incompatibility of
blocked clause elimination with redundancy criterion. In the end, we were left
somewhat disappointed to learn that SAT techniques do not scale good enough to be used
as superposition simplification techniques. However, we learned that they are
very successful as preprocessing techniques.

### 2022-03-30 summary/summary.tex
summary

\looseness=-1
In the last decades, proof assistants were immeasurably useful in formally
proving validity of hard mathematical theories and correctness of
safety-critical software. Using these tools one can formally describe a problem
and produce machine-checkable proofs for statements about it. To make the
checking phase efficient and trustworthy, proof steps need to be of high
granularity. As a consequence, seemingly simple statements must be proven in
minute detail, making the use of proof assistants very tedious.

@@

The first part extends E to support a fragment of higher-order logic devoid of
$\lambda$-abstra\-ction, yielding a prover called Ehoh. Even though this extension
is arguably conservative, Ehoh shows stronger performance on proof assistant
benchmarks than E.

The second part concerns adding support for $\lambda$-abstraction. The most
important challenge we faced during this extension is that of higher-order
unification. At the time we started, state-of-the-art procedure
for full higher-order unification was developed in 1970s. We designed a new
procedure inspired by this one that enumerates less redundant unifiers, has many
modern optimizations built-in, and can easily be customized to trade its
completeness for performance. It is implemented in a
prototype prover Zipperposition, a less efficient, but more easily extendible
prover compared to E. Our evaluation shows that our
procedure substantially improves on state-of-the-art.

>>>>

\looseness=-1
In the last decades, proof assistants were immeasurably useful in formally
proving validity of hard mathematical theories and correctness of
safety-critical software. Using these tools one can formally describe a problem
and produce machine-checkable proofs for statements about it. To make the
checking phase efficient and trustworthy, proof steps need to be of fine
granularity. As a consequence, seemingly simple statements must be proven in
minute detail, making the use of proof assistants very tedious.

@@

The first part extends E to support a fragment of higher-order logic devoid of
$\lambda$-abstra\-ction, yielding a prover called Ehoh. Despite this extension is small in scope, 
Ehoh shows stronger performance on proof assistant
benchmarks than E.

The second part concerns adding support for $\lambda$-abstraction. The most
important challenge we faced during this extension is that of higher-order
unification. At the time we started, state-of-the-art procedure
for full higher-order unification was developed in 1970s. We designed a new
procedure inspired by this one that enumerates fewer redundant unifiers, has many
modern optimizations built-in, and can easily be customized to trade its
completeness for performance. It is implemented in a
prototype prover called Zipperposition, a less efficient, but more easily extendible
prover compared to E. Our evaluation shows that our
procedure substantially improves on the state of the art.
